\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-15 writing/manuscript.Rmd
added some refs

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. **but animals don't only respond to changes in mean**

@@

The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.

@@

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months. @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring. However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck] or fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008]. @merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.

>>>>

While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range. Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce. Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance. More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022]. Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020]. Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982]. Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs. **but animals don't only respond to changes in mean**

@@

The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero). The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.

@@

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months. @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring. However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019]. *changes in bird migration, check for changes in mammals*

* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
